The state learner \\ {-- a super learner for right-censored data}

### 2024-05-13 tex/jrss-version/state-learner-jrss-style.tex
last comments

We compare the state learner to two IPCW based super learners: The
first superlearner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
% ANDERS: describe that you use a single evaluation time point and say how many
% censored, event happen before it.
%
% THOMAS: Done below now, OK?
%
% | time | sim_setting | true_events | true_cens | at_risk |
% |------+-------------+-------------+-----------+---------|
% |   36 | original    |      24.619 |    61.853 |  25.774 |
% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
%
The IPCW super learners use the integrated Brier score up to a fixed time
horizon (36 months). The marginal risk of the event before this time horizon is
25\%. Under the ``dependent censoring'', setting the marginal censoring
probability before this time horizon is 62\%; under the ``independent
censoring'' setting, the marginal censoring probability before this time horizon
is 39\%.

@@

The results are shown in Figure~\ref{fig:zelefski}. We see that for
most sample sizes, the state learner selected prediction models for
both censoring and outcome which have similar or higher IPA compared
to the prediction models selected by \texttt{survSL}.
% ANDERS: it is a bit lazy to write "for most sample sizes" :) it is
% slightly annoying that the y-axis scale changes in the 4
% figures. would that be better with Brier instead of IPA?
%
% THOMAS: No, it is worse with Brier score I seem to recall. I have fixed to
% y-axis horizontally now, it works well for the outcome model but poorly for
% the censoring. Let me know what you like the best.

\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
  \caption[]{For the risk prediction models of both the outcome and the
    censoring probability provided by each of the super learners, the IPA is
    plotted against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  % ANDERS: would be good to know how how many trainings set individuals were
  % right censored before the evaluation time point of the IPA. maybe of
  % interest how you chose the evaluation time point.
  %
  % THOMAS: Good point, but I think that the time of evaluation did not matter
  % too much. Of course, if you pick a very early time point where nothing has
  % happened, there is almost no difference in performance.
  \label{fig:zelefski}
\end{figure}

@@

% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer:
% in this section we re-analyse the ... Det skal ogsaa vaere klart i starten af
% sektion af der kommer noget om average treatment effect.
%
% THOMAS: Yes, klart. Fungerer det her:
In this section we re-analyze the prostate cancer data analyzed by
\cite{kattan2000pretreatment}, which we introduced in
Section~\ref{sec:numer-exper}. The data consists of 1,042 patients and includes
a competing event in the form of death without tumor recurrence. We use the
state learner to rank learners of the cause-specific cumulative hazard
functions. We then use the highest ranked combination of learners and the
estimation strategy outlined in Section~\ref{sec:targeted-learning} to estimate
the average treatment effect of hormone therapy on death and tumor recurrence.

We consider death without tumor recurrence and recurrence of tumor as competing
risks. We include the fives learner \texttt{KM}, \texttt{Cox strata CS},
\texttt{Lasso}, \texttt{Elastic}, and \texttt{RF} which are described in
Table~\ref{tab:zel-library}. We use the same library of learners to learn
\( \Lambda_1 \), \( \Lambda_2 \), and $\Gamma$. In this case, \( \Lambda_1 \)
denotes the cause-specific cumulative hazard function of tumor recurrence, and
\( \Lambda_2 \) denotes the cause-specific cumulative hazard function of death
without tumor recurrence.

@@

We have proposed a new super learner that can be used with
right-censored data and competing events. Compared to existing
IPCW-based methods, the advantage of the state learner is that it does
not depend on a pre-specified estimator of the censoring distribution,
but selects one automatically based on a library of learners for the
censoring distribution. Furthermore, the state learner neither
requires that \( \Lambda_j \) can be written as an integral with respect to
Lebesgue measure nor assumes a
% ANDERS: "density of \Lambda_j" er uklar hvordan er denne density defineret?
%
% THOMAS: Better now?
(semi-)parametric form for $\Lambda_j$. In the remainder of this
section we point to some limitations of our proposal and discuss
avenues for further research.

A major advantage of the state learner is that the performance of each
combination of learners can be estimated without (semi)parametric
assumptions or additional nuisance parameters. The drawback with this
approach is that we are rarely interested in features of the observed
data distribution when the data are right-censored: we are only
interested in the uncensored event time distribution. The finite
sample oracle inequality in Corollary~\ref{cor:oracle-prop} concerns
the function \( F \), which is a feature of \( P \in \mathcal{P} \),
while what we are typically interested in is \( \Lambda_j \) or
\( S \), which are features of \( Q \in \mathcal{Q} \). We emphasise
that while the state learner provides us with estimates of
\( \Lambda_j \) and $\Gamma$ based on libraries \( \mathcal{A}_j \)
and \( \mathcal{B} \), the performance of these learners is not
assessed directly for their respective target parameters, but only
indirectly via the performance of \( F \).
% ANDERS: er du enig med min omformulering af den sidste saetning?
%
% THOMAS: Yes, super godt. 
For settings without competing risks, our numerical studies suggest
that measuring the performance of \( F \) also leads to good
performance for estimation of \( S \). However, further research on
this issue, both numerical and theoretical, is warranted.
% ANDERS: ved ikke helt hvad du taenker man kan goer i fremtiden men synes
% egentig at den sidste saetning kunne godt slettes. :)
%
% THOMAS: Forstår ikke helt din kommentar. Hvilken sætning, den ovenover din
% kommentar eller den nedenunder? :)

% In the context of targeted learning, a drawback of the state learner is that the
% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
% lost because we only use a single nuisance parameter estimator. As defined here,
% however, the state learner is build using libraries for the conditional
% cause-specific cumulative hazard functions, so some doubly robustness might be
% preserved.

Our proposed super learner can be implemented with a broad library of learners
and using existing software.
% For instance the \texttt{R}-package
% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
Furthermore, while
the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
evaluate the performance of each learner we need to perform
\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
integrated Brier score in each hold-out sample, one for each combination of the
fitted models, but these operations are often negligible compared to fitting the
models. Hence the state learner is essentially not more computationally demanding
than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
and $\Gamma$ separately. While our proposal is based on constructing the library
\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
$\Gamma$, it would also be of interest to consider learners that estimate
\( F \) directly.
% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would
% not provide risk prediction models or would it? if yes you could say it
%
% THOMAS: I don't think that estimators for F exists -- that is what Vi is
% working on, right? Given F we can always obtain \Lambda_1 and \Lambda_2
% (equation 8 or eq:7 in tex) and then we can get predictions from these.

>>>>

We compare the state learner to two IPCW based super learners: The
first superlearner, called IPCW(Cox), uses a Cox model with main
effects to estimate the censoring probabilities, while the second
super learner, called IPCW(KM), uses the Kaplan-Meier estimator to
estimate the censoring probabilities. The Cox model for the censoring
distribution is correctly specified in both simulation settings while
the Kaplan Meier estimator only estimates the censoring model
correctly in the simulation setting where censoring is
independent. Both IPCW super learners are fitted using the
\texttt{R}-package \texttt{riskRegression}
\citep{Gerds_Ohlendorff_Ozenne_2023}.
%
% | time | sim_setting | true_events | true_cens | at_risk |
% |------+-------------+-------------+-----------+---------|
% |   36 | original    |      24.619 |    61.853 |  25.774 |
% |   36 | indep_cens  |      24.674 |    38.740 |  46.141 |
%
The IPCW super learners use the integrated Brier score up to a fixed time
horizon (36 months). The marginal risk of the event before this time horizon is
\(\sim 24.6\)\%. Under the ``dependent censoring'' setting the marginal censoring
probability before the time horizon is \(\sim 61.9\)\%. Under the ``independent
censoring'' setting the marginal censoring probability before this time horizon
is \(\sim 38.7\)\%.

@@

The results are shown in Figure~\ref{fig:zelefski}. We see that for
most sample sizes, the state learner selected prediction models for
both censoring and outcome which have similar or higher IPA compared
to the prediction models selected by \texttt{survSL}.
% ANDERS: Divide Fig 3 into two figures. keep y-axis fix for Outcome model panels but use different y-axes for  Censoring model.
\begin{figure}
  \centering %
  \includeFigCond[width=1\linewidth]{experiment-fig-sl-survSL.pdf}
  \caption[]{For the risk prediction models of both the outcome and the
    censoring probability provided by each of the super learners, the IPA at the fixed time horizon is
    plotted against sample size. The results are averages across 1000 repetitions and
    the error bars are used to quantify the Monte Carlo uncertainty.}
  % ANDERS: add marginal censoring probability to panel titles
  \label{fig:zelefski}
\end{figure}

@@

% ANDERS: har brug for en gentle intro saetning som fortaeller hvad der kommer:
% in this section we re-analyse the ... Det skal ogsaa vaere klart i starten af
% sektion af der kommer noget om average treatment effect.
%
% THOMAS: Yes, klart. Fungerer det her:
In this section we use the prostate cancer data of
\cite{kattan2000pretreatment} to illustrate the use of the state
learner in the presence of competing risks. We have introduced the
data in Section~\ref{sec:numer-exper}. The data consists of 1,042
patients who are followed from start of followup until tumor
recurrence, death without tumor recurrence or end of followup
(censored) whatever came first. For the sole purpose of illustration,
we estimate the average treatment effect of hormone therapy on death
and tumor recurrence. To do this we adapt the estimation strategy of
Section~\ref{sec:targeted-learning} as follows.  We use the state
learner to rank libraries of learners for the cause-specific
cumulative hazard functions of tumor recurrence, death without tumor
recurrence, and censoring.  The libraries of learners include the five
learners described in Table~\ref{tab:zel-library}.  We use the same
set of learners to learn the cumulative hazard function of tumor
recurrence \( \Lambda_1 \), the cumulative hazard function of death
without tumor recurrence \( \Lambda_2 \), and the cumulative hazard
function of the conditional censoring distribution $\Gamma$. 
We then use the highest ranked combination
of learners and apply formula \ref{eq:one-step-comp-ate}.

@@

We have proposed a new super learner that can be used with
right-censored data and competing events. Compared to existing
IPCW-based methods, the advantage of the state learner is that it does
not depend on a pre-specified estimator of the censoring distribution,
but selects one automatically based on a library of learners for the
censoring distribution. Furthermore, the state learner neither
requires that \( \Lambda_j \) can be written as an integral with
respect to Lebesgue measure, nor assumes a (semi-)parametric form for
$\Lambda_j$. In the remainder of this section we point to some
limitations of our proposal and discuss avenues for further research.

A major advantage of the state learner is that the performance of each
combination of learners can be estimated without (semi)parametric
assumptions or additional nuisance parameters. The drawback with this
approach is that we are rarely interested in features of the observed
data distribution when the data are right-censored: we are only
interested in the uncensored event time distribution. The finite
sample oracle inequality in Corollary~\ref{cor:oracle-prop} concerns
the function \( F \), which is a feature of \( P \in \mathcal{P} \),
while what we are typically interested in is \( \Lambda_j \) or
\( S \), which are features of \( Q \in \mathcal{Q} \). We emphasise
that while the state learner provides us with estimates of
\( \Lambda_j \) and $\Gamma$ based on libraries \( \mathcal{A}_j \)
and \( \mathcal{B} \), the performance of these learners is not
assessed directly for their respective target parameters, but only
indirectly via the performance of \( F \).
For settings without competing risks, our numerical studies suggest
that measuring the performance of \( F \) also leads to good
performance for estimation of \( S \).
% In the context of targeted learning, a drawback of the state learner is that the
% doubly robustness property defined in equation~(\ref{eq:dr-property}) seem to be
% lost because we only use a single nuisance parameter estimator. As defined here,
% however, the state learner is build using libraries for the conditional
% cause-specific cumulative hazard functions, so some doubly robustness might be
% preserved.

Our proposed super learner can be implemented with a broad library of learners
and using existing software.
% For instance the \texttt{R}-package
% \texttt{riskRegression} \citep{Gerds_Ohlendorff_Ozenne_2023} and the \texttt{R}-package by westling
Furthermore, while
the library \( \mathcal{F}(\mathcal{A}_1,\mathcal{A}_2,\mathcal{B}) \) consists
of \( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many learners, we only need to fit
\( |\mathcal{A}_1| +|\mathcal{A}_2| + |\mathcal{B}| \) many learners in each fold. To
evaluate the performance of each learner we need to perform
\( |\mathcal{A}_1||\mathcal{A}_2||\mathcal{B}| \) many operations to calculate the
integrated Brier score in each hold-out sample, one for each combination of the
fitted models, but these operations are often negligible compared to fitting the
models. Hence the state learner is essentially not more computationally demanding
than any procedure that uses super learning to learn $\Lambda_1$, $\Lambda_2$,
and $\Gamma$ separately. While our proposal is based on constructing the library
\( \mathcal{F} \) from libraries for learning \( \Lambda_1 \), $\Lambda_2$, and
$\Gamma$, it could also be of interest to consider learners that estimate
\( F \) directly.
% ANDERS: how would one estimate F? using HAL? or RF? but such a learner would
% not provide risk prediction models or would it? if yes you could say it
%
% THOMAS: I don't think that estimators for F exists -- that is what Vi is
% working on, right? Given F we can always obtain \Lambda_1 and \Lambda_2
% (equation 8 or eq:7 in tex) and then we can get predictions from these.
